Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2017-12-10 docs/ACM/paper.tex
Guido's pass 2

Annotation systems have risen to a place of prominence as a simple and
effective means to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10},
OpenSs~\cite{MeenderinckJ11}, Cilk++~\cite{Leiserson09} and
Tareador~\cite{Ayguade15}.
Annotations work as a meta-language: they let developers grant parallel
semantics to syntax originally written to execute sequentially.
%The compiler-related literature contains a generous load of success stories
%describing the use of annotation systems to parallelize programs.
Combined with modern accelerators, they have led to substantial performance
gains~\cite{Bertolli14,Mendonca17,Poesia17,Reyes12,Wienke12}.
Nevertheless, although convenient, the use of annotations is not straightforward, and
still lacks supporting tools.
%replaced:
%Yet, even if convenient,

Annotations such as OpenMP or OpenACC still require developers to
worry about typical hassles of the parallel world, such as race conditions
and deadlocks.
Moreover, because they are used in tandem with imperative programming languages,
such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{Amini12,Guelton12,Mendonca16,Pingali11,Nugteren14}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Thus, task parallelism remains still uncharted land in what concerns automatic
annotation.
Such fact is unfortunate, since much of the power of current annotation
systems lays on their ability to create tasks~\cite{Ayguade09}.
This ability brings said systems closer to irregular programs such as those
that implement graphs and worklist algorithms \`{a} la
Pingali~\cite{Ben-Nun17,Kulkarni11,Pingali11}.
The purpose of this work is to address this omission.

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
replicated in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

The major benefit provided by \Taskminer is, essentially, performance and readability.
\Taskminer{} receives as input a C program, and produces, as output, a C
program annotated with OpenMP directives that enclose potential tasks.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14} or Bots~\cite{Duran09}
benchmark suites, are large and complex; thus, their manual annotation is a
time consuming and error prone task.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
their sequential --unannotated-- versions, as we report in Section~\ref{sec:eval}.

@@

The use of OpenMP pragmas as the concrete means to explore, without human
intervention, task parallelism in programs presents difficulties and
advantages.
Such challenges and benefits have guided the design and implementation of the
techniques that we discuss in Section~\ref{sec:sol}.
In the rest of this section, we discuss them.

@@

The main advantage of using OpenMP to split a program into parallel tasks is
its runtime.
This runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its correctness~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}.
An OpenMP task descriptor contains the task's status (running, waiting) and a
list of memory addresses the task registered as input arguments.

The process of checking for dependencies can be as simple as checking if an
incoming task accesses a memory address already in use by a task in flight.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a topological ordering of this
graph~\cite{Planas15}.

@@

This runtime support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated automatically by the
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, through runtime checks.
Therefore, the ability to disambiguate memory regions statically is not
essential for correctness, although it is important for efficiency reasons,
as we will mention in Challenge~\ref{ch:Regions}.

>>>>

Annotation systems have risen to a place of prominence as a simple and
effective means to write parallel programs.
Examples of such systems include OpenMP~\cite{JaegerCP15},
OpenACC~\cite{OpenACC20}, OpenHMPP~\cite{Andion14}, OpenMPC~\cite{Lee10},
OpenSs~\cite{MeenderinckJ11}, Cilk++~\cite{Leiserson09} and
Tareador~\cite{Ayguade15}.
Annotations work as a meta-language: they let developers grant parallel
semantics to syntax originally written to execute sequentially.
%The compiler-related literature contains a generous load of success stories
%describing the use of annotation systems to parallelize programs.
\guido{What do you mean by modern accelerators? Not clear.}
Combined with modern accelerators, they have led to substantial performance
gains~\cite{Bertolli14,Mendonca17,Poesia17,Reyes12,Wienke12}.
\guido{Suppport what?}
Nevertheless, although convenient, the use of annotations is not straightforward, and
still lacks supporting tools \new{that can help programmers check if the annotation is
correct or effective}.
%replaced:
%Yet, even if convenient,

Annotations such as OpenMP or OpenACC still require developers to
worry about typical hassles of the parallel world, such as race conditions
and deadlocks.
Moreover, because they are used in tandem with imperative programming languages,
such troubles are only made worse by pointer aliasing.
There exist tools that insert automatic annotations in
programs~\cite{Amini12,Guelton12,Mendonca16,Pingali11,Nugteren14}.
All these technologies explore data-parallelism -- the possibility of running
the same computation independently on different data.
Thus, task parallelism remains still uncharted land in what concerns automatic
annotation. Such fact is unfortunate, since much of the power of current annotation
systems lays on their ability to create tasks~\cite{Ayguade09}.
\guido{Why the ability to create tasks makes such programs similar to irregular code?
Which features of irregular code are resposible for such similarities? Not clear to the 
reader}
This ability brings said systems closer to irregular programs such as those
that implement graphs and worklist algorithms \`{a} la
Pingali~\cite{Ben-Nun17,Kulkarni11,Pingali11}.
The purpose of this work is to address this omission.

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
\rmv{replicated} \new{privatized}  in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

\guido{How about saving programmer's time by doing automatic annotation?}
The major \ed{benefits provided by \Taskminer are, essentially, performance and readability.}
\Taskminer{} receives as input a C program, and produces, as output, a C
program annotated with OpenMP directives that enclose potential tasks.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14} or Bots~\cite{Duran09}
benchmark suites, are large and complex; thus, their manual annotation is a
time consuming and error prone task.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
\guido{Revisit later if necessary}
their sequential --unannotated-- versions, as we report in Section~\ref{sec:eval}.

@@

The use of OpenMP pragmas as the concrete means to explore, without human
intervention, task parallelism in programs presents difficulties and
advantages. 
Such challenges and benefits have guided the design and implementation of the
techniques \rmv{that we discuss} \new{described in} ection~\ref{sec:sol}.
\guido{The techniques or the advantages/difficulties? Ambiguous.} 
In the rest of this section \ed{we discuss such advantage and difficulties.}

@@

\guido{The runtime is a consequence; the main advantage is the possibility of leaving to a
runtime the task of handling dependences dynamically.} 
The main advantage of using OpenMP \new{annotations} to split a program into parallel tasks is
\new{the ability  of leaving the job of  handling dependences to a runtime.}
\new{The OpenMP runtime maintains a task dependence graph which dynamically exposes more parallelism 
opportunities than those resulting from a conservative  compile-time  analysis.}
\new{In particular, the} runtime lets us circumvent the shortcomings that pointer aliasing
impose on the automatic parallelization of code.
Aliasing -- the possibility of two pointers dereferencing the same
memory region -- either prevents parallelism altogether, or forces compilers to
resort to complex runtime checks to ensure its correctness~\cite{Alves15,Rus02}.
The OpenMP runtime shields us from this problem, because
it already checks for dependencies among tasks, and dispatches them in a
correct order~\cite{LaGrone11}. \ed{To achieve that, OpenMP runtime uses task descriptors that contain}
\guido{The descriptor does not keep task status; this is done in the runtime.}
\rmv{task's status (running, waiting)} \ed{a task identifier and  a list of memory addresses corresponding to the task input /output arguments.}

The process of checking for dependencies can be as simple as checking if an
incoming task accesses a memory address already in use by \ed{an in-flight task}.
It can also be complex, involving checks over long ranges of addresses
(in case the runtime system supports dependence checking for memory ranges).
This process can also include memory labelling and renaming (in case the
runtime system is able to remove false dependencies).
Regardless of its capacity -- which is not the same across every OpenMP
implementation -- the runtime system will represent dependencies using a task
dependence graph (TDG)~\cite{Duran08}.
In this directed acyclic graph, nodes denote tasks and edges represent
dependences between them.
Tasks are dispatched for execution according to a \new{dynamic} topological ordering of this
graph~\cite{Planas15}.

@@

\guido{Now it is clear the relevance for irregular programs.}
\ed{The OpenMP runtime} support allows the parallelization of irregular applications,
such as programs that traverse data-structures formed by a mesh of pointers.
\guido{A litte bit more details: }
\new{In such programs, control construcs (e.g. {\tt if} statement) make the execution 
of certain pieces of code  dependent on the program input. When using 
task parallelism the runtime can capture such dependences. This is not possible 
otherwise, since compile-time dependence analyzers must be conservative and eventually
skip some parallelization opportunity.}

As an example, Figure~\ref{fig:ex_book_filter} shows an application that
finds patterns in lines of a book.
The book is given as an array of pointers; each pointer leads to a
string representing a --potentially-- different line.
The natural parallelization of this program consists in firing off a task
to process each line.
Figure~\ref{fig:ex_book_filter} has been annotated automatically by \rmv{the}
\Taskminer.
The OpenMP execution environment ensures the correct scheduling of the tasks
created at lines 9 and 11, \ed{by}  \guido{This is vague:} \rmv{runtime checks.} 
\new{assuring that the annotated input dependencies {\tt line} and {\tt pattern}  are 
respected at runtime.}
\guido{Not clear. Need more details here. Adding:} \new{Pointer-based dependencies
can address various memory regions. Hence, in order to enable the automatic
synthesis of task annotations a system must be able to precisely define which 
memory region a pointer dependency is pointing to.}
\guido{Not clear why it is not important for correctness.}
Therefore, the ability to disambiguate memory regions statically is not
essential for correctness, although it is important for efficiency reasons,
as we will mention in Challenge~\ref{ch:Regions}.
